Model-free and Bayesian Neural Network Model-based Deep Reinforcement Learning for Particle Accelerator Control Demonstrated on the FERMI FEL

### 2020-12-17 main.tex
- paper changes

	\title{Model-Based and Model-Free Deep Reinforcement Learning in Accelerator Operation with Application to FERMI FEL}

	\author{Simon Hirlaender}\email{simon.hirlaender@sbg.ac.at}
	%\thanks{These two authors contributed equally}
	% \affiliation{University of Malta}
	\affiliation{IDA Lab, University of Salzburg, 5020 Salzburg, Austria}
	\author{Niky Bruchon}%, Gianfranco Fenu, Felice Andrea Pellegrino, Erica Salvato}
	\affiliation{University of Trieste, Trieste, Italy}
	% \author{Gianluca Valentino}
	% \affiliation{University of Malta, Msida, MT}
	% \author{Marco Lonza, Giulio Gaio}
	% \affiliation{Elettra Sincrotrone Trieste, Trieste, IT}
	% \author{Verena Kain}
	% %\thanks{These two authors contributed equally}

	\date{\today}% It is always \today, today,
	%  but any date may be explicitly specified

	\begin{abstract}
		Reinforcement learning holds tremendous promise in accelerator controls, and the principal goal of this paper is to show how reinforcement learning can be utilised on an operational level on accelerator physics problems.
		We compare a model-based reinforcement to a model-free reinforcement learning method applied to the FERMI FEL system.  In terms of representational power and sample-efficiency, the model-based approach is overtaking, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the techniques were adapted to show increased noise robustness as omnipresent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands to consider several control problems simultaneously \cite{Kain2020,Scheinker2018}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	 Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL in real world applications is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}. A second critical aspect to be considered is the robustness of the training in a noisy data limited environment.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at \emph{Elettra Sincrotrone Trieste}.\\
	 We successfully applied two different algorithm classes, a highly sample-efficient variant of continuous Q-learning (\cref{ss:Normalized advantage function}) and a newly proposed 
	  model-based algorithm, which we call \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), to the FERMI FEL problem.\\ The methods reveal different characteristics addressing the stated critical aspects. They are generally applicable to problems showing a sample-complexity as in common accelerator control problems.
	 \subsection{An Overview of the Main Results}
	 We demonstrate optimization techniques that simultaneously adjust 4 parameters, the tilt and incline of 2 mirrors that control the trajectory and overlap of the seed laser in the FERMI FEL relative to the beam in the undulator. We show that our algorithms require significantly less than 1000 training sets, after which they are able to maximize the FERMI FEL’s output intensity to $>$ that 95\% of the maximum within only $\sim 3$ to $\sim 5$ steps depending on which algorithm is used. Also we show the operational applicability of purely model-based reinforcement learning in an accelerator control problem.\\
	  Furthermore, our algorithms can perform real-time learning based on data and therefore can be re-run and re-trained for time-varying systems.\\
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Discussion, outlook and summary
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

%	At FERMI several beam-based feedback systems are deployed to control the beams trajectories shot to shot with the main purpose of guaranteeing a steady intensity of the light radiation. 
In the last years, various approaches have been studied to investigate their applicability in tuning operations. \\
	A free-electron laser is a fourth-generation light source where the lasing medium consists of a very-high-speed electron moving freely through a magnetic structure. By using an external seeding source, the FERMI FEL has several advantages relative to standard FEL approaches in terms of increased stability in pulse and photon energy, reduced size, improved longitudinal coherence, more control on the pulse structure, and improved temporal synchronization with external timing systems.\\

@@

%	The most critical parameters in a seeded free-electron laser are the temporal and transverse overlap of the electron and laser beams in the modulator magnetic section. 

@@

	Two mirrors, TT1 and TT2, are used to control the trajectory of the laser by tilting and inclining, which gives a total of four degrees of freedom (DOF). In turn, the laser overlaps with the electron beam between the two screens, CCD1 and CCD2. Lastly, the monitor measures the intensity, $I_0$.\\

@@

	and is limited by an $ |\ba_t|\leq \ba_\text{max}=\frac{1}{12}$.
	The training is done in episodes with a uniformly\footnote{in the normalized state-space} randomized initial state and a specific maximal length of taken steps afterwards - the horizon. For reason discussed later, the horizon varies in the applications between either 10 or 500 steps. An early termination happens if a specific threshold is obtained, which lies in all cases at 95\% of the maximal achievable intensity of the system. In our experiments the maximal intensity was determined by a manual expert optimization. \\
	 During the training phase exploration and the training of the function approximator takes place.
	After a defined number of training episodes, a verification of 50 or 100 episodes is performed. In the second phase, the RL only applies the learned policy.

@@

	\section{Deep Reinforcement learning}
    Most commonly two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithm learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, there applications is mainly restricted to simulated environments due to their low sample-efficiency. Yet putting constrains on the algorithms, as discussed in \cref{s:Model free reinforcement learning}, partly cures the sample inefficiency turning them into applicable tools for real world problems.\\
      On the other hand MBRL showing a much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample-efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	Goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
	p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
	\label{eq:trajectory_distribution}
	\end{align}

@@

	\subsection{Model free reinforcement learning}\label{s:Model free reinforcement learning}

@@

	 In the second case we speak about Q-learning, which belongs to approximate dynamic programming. One advantage is the possible usage of use previously stored data as we use it in an \emph{off-policy} fashion, which generally needs less data to be trained than \emph{on-policy} methods. A combination of both is also used, known as actor-critic methods \cite{Szepesvari2010,Lillicrap2015,Silver2014}.\\
	 As already adduced, MFRL algorithms learn directly from interactions with the system. An exhaustive overview can be found e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm \cite{Gu2016} in some detail, as it has good characteristics for accelerator control problems. It is highly sample-efficiency  and its representational power is sufficient for most common cases\cite{Kain2020,Hirlaender2020a}. Modifications to increase the stability are subject of \cref{ss:Normalized advantage function}.

@@

	 Our implementation has a high sample-efficiency and noise resistivity. We use it as the baseline for the FERMI-FEL control problem, as it yields good results. Additional changes to the original proposal \cite{Gu2016} and a previous implementation used for accelerator control using to a prioritized replay buffer \cite{Hirlaender2020a} are discussed in \cref{appendix:naf2}. We refer to our modified implementation as \emph{NAF2} and the implementation close to the original, as in \cite{Gu2016}, as \emph{NAF}.

@@

	A nice overview of state of the art MBRL methods is provided in \cite{Wang2019}. Suitable methods are \cite{Gal2016,6654139}, based on Gaussian processes and Bayesian Neural Networks and back-propagation in the dynamics model or \cite{Chua2018,Wang2019a}, using highly efficient sample based methods (e.g. the cross entropy method \cite{Boer2005}). We focus on model-based data generation.\\
		In MFRL the dynamics $\transitions$ of an environment is learned implicitly, while in model-based reinforcement learning MBRL $\transitions$ is learned explicitly and approximated as $\hat \transitions$. In addition to the next state, the reward is included in the model \footnote{In some applications $\reward$ can be deduced from the current state or the state change. In such a situation the modelling can be simplified to model directly $\hat\transitions$}:

@@

	  After a defined number of training steps, called epoch, the controller is validated on each model of the ensemble, individually. If there is no improvement in a number smaller than the total number of the models, the controller might be tested on the real environment for several episodes. In this way 'over-fitting' on a wrong model is avoided, as discussed in \cite{Kurutach2018}. The training is stopped if the controller produces satisfying results on the real environment. Otherwise, a batch of new data is collected and the model is improved and consecutively a new epoch starts.

@@

The prior is controlled by the initialisation of the weights of the network. It defines how large the variation of the approximated function is expected and were randomized uniformly in a small interval $[-\Delta, \Delta]$. The weights of the last layer were normalised by the number of its nodes. For the moment, only homoskedastic Gaussian errors are considered with a standard deviation is $\sigma_\epsilon$ is also respected in a regularisation term \cref{appendix:The impact of noise} and significantly improves the training in the presence of noise, as shown in \cref{fig:comparsion_noise_ae_dyna}. \\

@@

To decide which MFRL algorithm to use, two already mentioned main algorithm classes are considered  (\cref{s:Model free reinforcement learning}): \emph{on-} and  \emph{off-policy} algorithms \cite{Sutton2018}. \emph{On-policy} algorithm show a stabler convergence in general, while \emph{off-policy} algorithms have the advantage that the data buffer can be filled with real data.\\
The \emph{on-policy} algorithm trust region policy optimization (TRPO) \cite{Schulman2015} provides the theoretical guarantee of a monotonic improvement of the policy. We use this algorithm in experiments labelled as \emph{ME-TRPO}, which stands for model-ensemble TRPO as proposed in \cite{Kurutach2018}. As reported in \cite{Kurutach2018}, using proximal policy optimisation \cite{Schulman2017}, another prominent \emph{on-policy} algorithm, reveals minor performance, which was confirmed in our tests. \\
An attractive off-policy algorithm is the soft-actor-critic (\emph{SAC}) algorithm. \cite{fujimoto2018addressing,Hill2018}. \emph{SAC} not only tries to maximise \cref{eq:cumulative_reward} but also simultaneously is regularised to maximise the entropy in the action space \cite{Haarnoja2018a}. In this way, exploration is encouraged to avoid getting stuck in a local optimum and a good trade-off between exploration and exploitation is achieved. We refer to experiments using the \emph{SAC} as \emph{AE-DYNA-SAC}. 
For \emph{AE-DYNA-SAC} the controller was reset after each retraining of the dynamics model, to profit from its exploration features, while for \emph{ME-TRPO} the policy was constantly improved to exploit its monotony.

@@

One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance the original implementation of the model ensemble \emph{TRPO} (\emph{ME-TRPO}) algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward (as used e.g. in \cite{kidambi2020morel}) and showed constant but too slow improvement for our online-training.

@@

	The main purpose was to try the newly implemented algorithms on a real system to evaluate their operational feasibility. 

@@

		\caption{The training of the \emph{NAF2} on the FERMI FEL using a single network (dashed) and a double network (solid).}

@@

		\caption{The training of the \emph{NAF2} on the FERMI FEL using a double network.}

@@

	\caption{The training of the \emph{NAF2} on the FERMI FEL using a single network.}

@@

		\caption{The verification of the \emph{NAF2} on the FERMI FEL using a single network (dashed) and a double network (solid).}

@@

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/FERMI_all_experiments_NAF_convergence.pdf}
		\caption{The training metrics of the \emph{AE-DYNA-SAC} on the FERMI FEL using a single network (dashed) and a double network (solid). The Bellman error (\cref{eq:minimize_bellmann_optimality}) and the state-value function (\cref{eq:state-value-function}) are shown. }
		\label{fig:NAF_convergence}
	\end{figure}
	In total, four tests were carried out, two using a single network and two using the double network architecture. 
	\Cref{fig:NAF_training} displays the results, averaged over the two tests \footnote{Considering the training is stochastic with random initial conditions, by chance, the training of the double network variant started above the threshold in both cases.}. A training of 100 episodes was accomplished. The number of iterations per episode is plotted, including the cumulative number of steps. In all episodes during the training the 95\% intensity reward threshold was surpassed, and therewith they were successfully finished. The evolution of the states is provided in \cref{fig:NAF_evolution_single,fig:NAF_evolution_double} for the single and double network. At the boundary of the domain, it takes time to learn, since many actions are mapped onto the same state\footnote{The states are clipped to $[0,1]$.}. The behaviour indicates that the double network converges faster and spends less time at the boundary.\\
	In the verification of 100 episodes, \cref{fig:NAF_verification}, both algorithms show similar performance, while the double network needed less training steps, and reveals a more stable overall performance. Additionally, the convergence metrics of the two algorithms is plotted in \cref{fig:NAF_convergence} against the number of training steps. The blue curves show the Bellmann error \cref{eq:minimize_bellmann_optimality}, which is comparable in both cases. The state-value function $V$ (\cref{eq:state-value-function}), which is a direct output of the neural net (\cref{eq:state-action-value-approxiation}), converges to a reasonable value for the double network within the shown 700 steps, whereas the single network seems to overestimate the value. In the single-network case, convergence is reached after around 1400 steps. In the plot $V$ was averaged over 200 randomly selected states.
	\subsection{MBRL tests}
	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/AE-DYNA_observables.pdf}
		\caption{The training of the \emph{AE-DYNA-SAC} on the FERMI FEL.}
		\label{fig:AE-DYNA_observables}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/Verification_DYNA_all_episodes.pdf}
		\caption{The verification of the \emph{ME-TRPO} and the \emph{AE-DYNA-SAC} on the FERMI FEL.}
		\label{fig:AE-DYNA_verification}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/ME-TRPO_observables.pdf}
		\caption{The training of the ME-TRPO on the FERMI FEL.}
		\label{fig:ME-TRPO_observables}
	\end{figure}

@@

	\caption{The verification of the \emph{ME-TRPO} and the \emph{AE-DYNA-SAC} on the FERMI FEL.}

@@

>>>>

	\title{Model-free and Bayesian Neural Network Model-based Deep Reinforcement Learning for Particle Accelerator Control Demonstrated on the FERMI FEL}

	\author{Simon Hirlaender}\email{simon.hirlaender@sbg.ac.at}
	%\thanks{These two authors contributed equally}

	\affiliation{IDA Lab, University of Salzburg, 5020 Salzburg, Austria}
	\author{Niky Bruchon}
	\affiliation{University of Trieste, Trieste, Italy}	
	\date{\today}% It is always \today, today,

	\begin{abstract}
		Reinforcement learning holds tremendous promise in accelerator controls. The primary goal of this paper is to show how this approach can be utilised on an operational level on accelerator physics problems. Despite the success of model-free reinforcement learning in several domains, sample-efficiency still is a bottle-neck, which might be encompassed by model-based methods.
		We compare well-suited purely model-based to model-free reinforcement learning applied to the intensity optimisation on the FERMI FEL system. We find that the model-based approach demonstrates higher representational power and sample-efficiency, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the algorithms were implemented in a way, which presents increased noise robustness as omnipresent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands the consideration of several control problems simultaneously \cite{Kain2020,Scheinker2018}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020,John2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL in real-world applications is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}. A second critical aspect to be considered is the robustness of the training in a noisy data-limited environment.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at \emph{Elettra Sincrotrone Trieste}.\\
	We successfully applied two different algorithm classes, a highly sample-efficient variant of continuous Q-learning (\cref{ss:Normalized advantage function}) and a newly proposed model-based algorithm, which we call \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), to the FERMI FEL problem.\\ The methods reveal different characteristics addressing the stated critical aspects. They are generally applicable to problems showing a sample-complexity as in common accelerator control problems.
	\subsection{An Overview of the Main Results}
	We demonstrate optimisation techniques that simultaneously adjust four parameters, the tilt and incline of two mirrors that control the trajectory and overlap of the seed laser in the FERMI FEL relative to the beam in the undulator. 
%	We show that our algorithms require significantly less than 1000 training sets, after which they can maximise the FERMI FEL's output intensity to $>$ that 95\% of the maximum within only $\sim 3$ to $\sim 5$ steps depending on which algorithm is used.
	By making use of model ensembles and Bayesian neural networks our algorithms exhibit a high sample-efficiency (learning using $<1000$ data sets), a low model-bias, and are extremely fast as they are able to maximize the FERMI FEL's output intensity to greater than 95\% of the maximum within only 3-5 steps.
	 Also, we show the operational applicability of purely model-based reinforcement learning in an accelerator control problem.\\
	Furthermore, our algorithms can perform real-time learning based on data and therefore, can be re-run and re-trained for time-varying systems.\\
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Discussion, outlook and summary
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

	%	At FERMI, several beam-based feedback systems are deployed to control the beams trajectories shot to shot with the primary purpose of guaranteeing a steady intensity of the light radiation. 
	In the last years, various approaches have been studied to investigate their applicability in machine tuning. \\
	A free-electron laser is a fourth-generation light source where the lasing medium consists of a very-high-speed electron moving freely through a magnetic structure. By using an external seeding source, the FERMI FEL has several advantages relative to standard FEL approaches in terms of increased stability in pulse and photon energy, reduced size, improved longitudinal coherence, more control on the pulse structure, and improved temporal synchronisation with external timing systems.\\

@@

	%	The most critical parameters in a seeded free-electron laser is the temporal and transverse overlap of the electron and laser beams in the modulator magnetic section. 

@@

	Two mirrors, TT1 and TT2, are used to control the trajectory of the laser by tilting and inclining, which gives a total of four degrees of freedom (DOF). In turn, the laser overlaps with the electron beam between the two removable screens, CCD1 and CCD2. Lastly, the monitor measures the intensity, $I_0$.\\

@@

	and is limited by an $ |\ba_t|\leq \ba_\text{max}=1/12$.
	The training is done in episodes with a uniformly\footnote{in the normalised state-space} randomised initial state and a specific maximal length of taken steps afterwards - the horizon. For a reason discussed later, the horizon varies in the applications between either 10 or 500 steps. An early termination happens if a specific threshold is obtained, which lies in all cases at 95\% of the maximal achievable intensity of the system. In our experiments, the maximal intensity was determined by manual expert optimisation. \\
	During the training phase exploration and the training of the function, approximator takes place.
	After a defined number of training episodes, verification of 50 or 100 episodes is performed. In the second phase, the RL only applies the learned policy.

@@

	\section{Deep Reinforcement learning}
	Most commonly, two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while the MBRL algorithm learns a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, their applications is mainly restricted to simulated environments due to their low sample-efficiency. However, putting constraints on the algorithms, as discussed in \cref{s:Model-free reinforcement learning}, partly cures the sample inefficiency, turning them into practical tools for real-world problems.\\
	On the other hand, MBRL showing a much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so-called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample-efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	Goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
		p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
		\label{eq:trajectory_distribution}
	\end{align}

@@

	\subsection{Model free reinforcement learning}\label{s:Model-free reinforcement learning}

@@

	In the second case, we speak about Q-learning, which belongs to approximate dynamic programming. One advantage is the possible usage of use previously stored data as we use it in an \emph{off-policy} fashion, which generally needs fewer data to be trained than \emph{on-policy} methods. A combination of both is also used, known as actor-critic methods \cite{Szepesvari2010,Lillicrap2015,Silver2014}.\\
	As already adduced, MFRL algorithms learn directly from interactions with the system. An exhaustive overview can be found, e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm \cite{Gu2016} in some detail, as it has good characteristics for accelerator control problems. It is highly sample-efficiency, and its representational power is sufficient for most common cases\cite{Kain2020,Hirlaender2020a}. Modifications to increase the stability are subject of \cref{ss:Normalized advantage function}.

@@

	Our implementation has high sample-efficiency and noise resistivity. We use it as the baseline for the FERMI-FEL control problem, as it yields good results. Additional changes to the original proposal \cite{Gu2016} and a previous implementation used for accelerator control using to a prioritized replay buffer \cite{Hirlaender2020a} are discussed in \cref{appendix:naf2}. We refer to our modified implementation as \emph{NAF2} and the implementation close to the original, as in \cite{Gu2016}, as \emph{NAF}.

@@

	An excellent overview of the state of the art MBRL methods is provided in \cite{Wang2019}. Suitable methods are \cite{Gal2016,6654139}, based on Gaussian processes and Bayesian Neural Networks and back-propagation in the dynamics model or \cite{Chua2018,Wang2019a}, using highly efficient sample-based methods (e.g. the cross entropy method \cite{Boer2005}). We focus on model-based data generation.\\
	In MFRL the dynamics $\transitions$ of an environment is learned implicitly, while in model-based reinforcement learning MBRL $\transitions$ is learned explicitly and approximated as $\hat \transitions$. In addition to the next state, the reward is included in the model \footnote{In some applications $\reward$ can be deduced from the current state or the state change. In such a situation, the modelling can be simplified to model directly $\hat\transitions$}:

@@

	After a defined number of training steps, called an epoch, the controller is validated on each model of the ensemble, individually. If there is no improvement in a number smaller than the total number of the models, the controller might be tested on the real environment for several episodes. In this way, 'over-fitting' on a wrong model is avoided, as discussed in \cite{Kurutach2018}. The training is stopped if the controller produces satisfying results on the real environment. Otherwise, a batch of new data is collected, and the model is improved, and consecutively a new epoch starts.

@@

	The prior is controlled by the initialisation of the weights of the network. It defines how large the variation of the approximated function is expected and they were randomised uniformly in an interval $[-\Delta, \Delta]$, where $|\Delta|\leq0.1$. The weights of the last layer were normalised by the number of its nodes. For the moment, only homoskedastic Gaussian errors are considered with a standard deviation $\sigma_\epsilon$. This is respected in a regularisation term (details see \cref{appendix:The impact of noise}) and significantly improves the training in the presence of noise, as shown in \cref{fig:comparsion_noise_ae_dyna}. \\

@@

	To decide which MFRL algorithm to use, two already mentioned main algorithm classes are considered  (\cref{s:Model-free reinforcement learning}): \emph{on-} and  \emph{off-policy} algorithms \cite{Sutton2018}. \emph{On-policy} algorithm show a stabler convergence in general, while \emph{off-policy} algorithms have the advantage that the data buffer can be filled with real data.\\
	The \emph{on-policy} algorithm trust region policy optimisation (TRPO) \cite{Schulman2015} provides the theoretical guarantee of a monotonic improvement of the policy. We use this algorithm in experiments labelled as \emph{ME-TRPO}, which stands for model-ensemble TRPO as proposed in \cite{Kurutach2018}. As reported in \cite{Kurutach2018}, using proximal policy optimisation \cite{Schulman2017}, another prominent \emph{on-policy} algorithm, reveals minor performance, which was confirmed in our tests. \\
	An attractive \emph{off-policy} algorithm is the soft-actor-critic (\emph{SAC}) \cite{fujimoto2018addressing,Hill2018}. \emph{SAC} not only tries to maximise \cref{eq:cumulative_reward} but also simultaneously is regularised to maximise the entropy in the action space \cite{Haarnoja2018a}. In this way, exploration is encouraged to avoid getting stuck in a local optimum, and a good trade-off between exploration and exploitation is achieved. We refer to experiments using the \emph{SAC} as \emph{AE-DYNA-SAC}. 
	For \emph{AE-DYNA-SAC} the controller was reset after each re-training of the dynamics model, to profit from its exploration features, while for \emph{ME-TRPO} the policy was improved continuously to exploit its monotony.

@@

	One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance the original implementation of the model ensemble \emph{TRPO} (\emph{ME-TRPO}) algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward (as used, e.g. in \cite{kidambi2020morel}) and showed constant but too slow improvement for our online-training.

@@

	The primary purpose was to try the newly implemented algorithms on a real system to evaluate their operational feasibility. 

@@

		\caption{The training of different variants of the \emph{NAF2} algorithm  on the FERMI FEL, averaged over two complete trainings (the standard-deviations are indicated by the shaded areas). The number of iterations (blue) shows the steps until the intensity is optimised, starting from a random initial position.}

@@

		\caption{The evolution of the states of the \emph{NAF2} algorithm on the FERMI FEL using a double network during the training.}

@@

		\caption{The evolution of the states of the \emph{NAF2} algorithm on the FERMI FEL using a single network during the training.}

@@

		\caption{The verification episodes of the variants of the trained model-free \emph{NAF2} algorithm on the FERMI FEL. The number of iterations (blue) shows the steps until the intensity is optimised, starting from a random initial position.}

@@

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/FERMI_all_experiments_NAF_convergence.pdf}
		\caption{The training metrics of the \emph{AE-DYNA-SAC} on the FERMI FEL using a single network (dashed) and a double network (solid). The Bellman error (\cref{eq:minimize_bellmann_optimality}) and the state-value function (\cref{eq:state-value-function}) are shown. }
		\label{fig:NAF_convergence}
	\end{figure}
	In total, four tests were carried out, two using a single network and two using the double network architecture. 
	\Cref{fig:NAF_training} displays the results, averaged over the two tests \footnote{Considering the training is stochastic with random initial conditions, by chance, the training of the double network variant started above the threshold in both cases.}. A training of 100 episodes was accomplished. The number of iterations per episode is plotted, including the cumulative number of steps. In all episodes, during the training, the 95\% intensity reward threshold was surpassed, and therewith they were successfully finished. The evolution of the states is provided in \cref{fig:NAF_evolution_single,fig:NAF_evolution_double} for the single and double network. At the boundary of the domain, it takes time to learn, since many actions are mapped onto the same state\footnote{The states are clipped to $[0,1]$.}. The behaviour indicates that the double network converges faster and spends less time at the boundary.\\
	In the verification of 100 episodes, \cref{fig:NAF_verification}, both algorithms show similar performance, while the double network needed less training steps, and reveals a more stable overall performance. Additionally, the convergence metrics of the two algorithms is plotted in \cref{fig:NAF_convergence} against the number of training steps. The blue curves show the Bellmann error \cref{eq:minimize_bellmann_optimality}, which is comparable in both cases. The state-value function $V$ (\cref{eq:state-value-function}), which is a direct output of the neural net (\cref{eq:state-action-value-approxiation}), converges to a reasonable value for the double network within the shown 700 steps, whereas the single network seems to overestimate the value. In the single-network case, convergence is reached after around 1400 steps. In the plot, $V$ was averaged over 200 randomly selected states.
	\subsection{MBRL tests}
	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/AE-DYNA_observables.pdf}
		\caption{The training observables of the \emph{AE-DYNA-SAC} on the FERMI FEL.}
		\label{fig:AE-DYNA_observables}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/Verification_DYNA_all_episodes.pdf}
		\caption{The verification episodes of the trained model-based methods: \emph{ME-TRPO} and \emph{AE-DYNA-SAC} on the FERMI FEL. The number of iterations (blue) shows the steps until the intensity is optimised, starting from a random initial position.}
		\label{fig:AE-DYNA_verification}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics*[width=0.5\textwidth]{Figures/ME-TRPO_observables.pdf}
		\caption{The training observables of the \emph{ME-TRPO} on the FERMI FEL.}
		\label{fig:ME-TRPO_observables}
	\end{figure}

@@

		\caption{The evolution of the states during the worst verification episodes of the trained \emph{ME-TRPO} and the \emph{AE-DYNA-SAC} on the FERMI FEL.}

@@

	\begin{figure}[h]
